### 2022-12-12 intro.Rnw
one-sided sceptical p-value

\begin{figure}[!htb]
<< "RPCB-graph", fig.height = 3.5 >>=
za <- qnorm(p = 0.975)
rpcb <- read.csv("data/rpcb.csv") |>
    mutate(id = paste0("(", paper, ", ", experiment, ", ", effect, ")"),
           zo = smdo/so,
           zr = smdr/sr,
           c = so^2/sr^2,
           d = smdr/smdo,
           po2 = 2*(1 - pnorm(q = abs(zo))),
           pr2 = 2*(1 - pnorm(q = abs(zr))),
           sm = 1/sqrt(1/so^2 + 1/sr^2),
           smdm = (smdo/so^2 + smdr/sr^2)*sm^2,
           pm2 = 2*(1 - pnorm(q = abs(smdm/sm))),
           Q = (smdo - smdr)^2/(so^2 + sr^2),
           pQ = pchisq(q = Q, df = 1, lower.tail = FALSE),
           rsig = factor(pr2 <= 0.05, levels = c(FALSE, TRUE),
                         labels = c("italic(p[r]) > 0.05",
                                    "italic(p[r]) <= 0.05")))
## compute sceptical p-value (TODO fix problems with NA)
ind <- !is.na(rpcb$zo) & !is.na(rpcb$zr) & !is.na(rpcb$c)
rpcb$ps2 <- NA
rpcb$ps2[ind] <- pSceptical(zo = rpcb$zo[ind], zr = rpcb$zr[ind], c = rpcb$c[ind],
                            alternative = "two.sided", type = "nominal")

@@

A detailed view for a subset of the data from the project is given in
Table~\ref{tab:RPCB}. The subset is chosen such that from every original study
one effect and one replication of that effect is included. For no single study
pair in the table are all commonly used criteria satisfied simultaneously. For
instance, the first two replications satisfy the significance criterion
($p_{o} < 0.05$ and $p_{r} < 0.05$), meta analytic significance ($p_{m} < 0.05$)
and the $Q$-test/prediction interval criterion ($p_{Q} > 0.05$), yet the
replication effect estimate is smaller than the original one ($d < 1$).
Similarly, there is no single study pair for which all criteria indicate
replication failure. For example, the last replication (\#20) is a clear failure
with respect to the relative effect estimate ($d < 1$) and the significance
criteria ($p_{o} > 0.05$ and $p_{r} > 0.05$), yet the $Q$-test does not indicate
evidence for inconsistency of the two effect estimates ($p_{Q} > 0.05$).

@@

\begin{landscape}
\begin{table}
  \centering
  \caption{Subset of results from the ``Reproducibility Project: Cancer
    Biology'' \citep{Errington2021}. (P, X, E) denotes effect number E from
    experiment number X, from original paper number P. Shown are original
    $\that_{o}$, replication $\that_{r}$, and pooled effect estimate $\that_{m}$
    with 95\% confidence intervals, variance ratio
    $c = \sigma^{2}_{o}/\sigma^{2}_{r}$, relative effect estimate
    $d = \that_{r}/\that_{o}$, original \textit{p}-value $p_{o}$, replication
    \textit{p}-value $p_{r}$, meta-analytic \textit{p}-value $p_{m}$, $Q$-test
    \textit{p}-value $p_{Q}$, and sceptical \textit{p}-value $p_{\mathrm{S}}$. A
    replication success (green) threshold of $\alpha = 5\%$ is used for the
    \textit{p}-values and $d_{\text{min}} = 1$ for the relative effect
    estimate.}
  \label{tab:RPCB}
\resizebox{1.02\linewidth}{!}{%
<< "RPCB-table2", results = "asis" >>=
## function to format CI
ciFormat. <- function(lower, upper, dollar = FALSE, digits = 2) {
    if (dollar) x <- "$" else x <- ""
    paste0("(", x, format(round(lower, digits), nsmall = digits), x,
           " to ", x, format(round(upper, digits), nsmall = digits), x, ")")
}
ciFormat <- Vectorize(FUN = ciFormat.)

@@

## randomly select a subset with one effect per paper
set.seed(7)
rpcbTable <- rpcb |>
    na.omit() |>
    group_by(paper) |>
    sample_n(size = 1) |>
    ungroup() |>
    arrange(pr2) |>
    mutate(smdoFormat = signif(smdo, 3),
           smdrFormat = signif(smdr, 3),
           smdmFormat = signif(smdm, 3),
           dFormat = tableFormat(x = d, outFun = dFormat,
                                 threshold = 1, col1 = "dark2orange",
                                 col2 = "dark2green"),
           cioL = smdo - za*so,
           cioU = smdo + za*so,
           cirL = smdr - za*sr,
           cirU = smdr + za*sr,
           cimL = smdm - za*sm,
           cimU = smdm + za*sm,,
           po2Format = tableFormat(x = po2),
           pr2Format = tableFormat(x = pr2),
           pm2Format = tableFormat(x = pm2),
           ps2Format = tableFormat(x = ps2),
           pQFormat = tableFormat(x = pQ),
           cioFormat = ciFormat(lower = cioL, upper = cioU, dollar = TRUE),
           cirFormat = ciFormat(lower = cirL, upper = cirU, dollar = TRUE),
           cimFormat = ciFormat(lower = cimL, upper = cimU), dollar = TRUE) |>
    select(id,
           smdoFormat, cioFormat,
           smdrFormat, cirFormat,
           smdmFormat, cimFormat,
           c, dFormat,
           po2Format, pr2Format, pm2Format, pQFormat, ps2Format
           )
rpcbXtable <- xtable(rpcbTable)
colnames(rpcbXtable) <- c("(P, X, E)",
                          "$\\that_o$", "(95\\% CI)",
                          "$\\that_r$", "(95\\% CI)",
                          "$\\that_m$", "(95\\% CI)",
                          "$c$",
                          "$d$",
                          "$p_o$",
                          "$p_r$",
                          "$p_m$",
                          "$p_Q$",
                          "$p_{\\text{S}}$"
                          )

@@

The sceptical $p$-value has several interesting properties \citep[Section
3]{Held2020}: First, it is always larger than the maximum of the original and
replication $p$-values ($p_{\text{S}} > \max\{p_{o}, p_{r}\}$), meaning that
both $p$-values have to be smaller than $\alpha$ such that replication success
at level $\alpha$ is possible. The sceptical $p$-value hence requires
\emph{both} studies to be sufficiently convincing on their own (in terms of
their $p$-values), similar to the significance criterion for replication
success. Second, if the $p$-values $p_{o}$ and $p_{r}$ remain fixed but the
relative effect estimate $d = \that_{r}/\that_{o}$ decreases, the sceptical
$p$-value increases, meaning that shrinkage of the replication effect estimate
is penalized ($p_{\text{S}} \uparrow 1$ as $d \downarrow 0$ for fixed $p_{o}$
and $p_{r}$). The sceptical $p$-value thus takes effect shrinkage into account,
unlike the significance criterion which can be achieved with any non-zero
replication effect estimate $\that_{r}$ provided the standard error $\sigma_{r}$
is small enough. This property is desirable in the replication setting as a
smaller replication effect estimate $\that_{r}$ may not be practically relevant
anymore, despite its statistical significance.

The results from the ``Reproducibility Project: Cancer Biology'' in
Table~\ref{tab:RPCB} illustrate these two properties. The third study fails to
achieve replication success at level $\alpha = 5\%$ with the sceptical
$p$-value, even though the replication study was highly convincing (the effect
estimate was almost five times as large as in the original study). Yet, as the
approach requires both studies to be convincing on their own -- and the original
study was not significant at the 5\% level -- replication success at this level
is impossible with the sceptical $p$-value. The second property is illustrated
by the fifth study. Here the original study was convincing and the replication
study also achieved significance. However, the effect estimate from the
replication was roughly 60\% smaller than the one from the original study, and
significance is merely achieved because the standard error was much smaller
(\ie{} around $1/\surd c \approx 1/2.8$ times smaller). The sceptical $p$-value
is therefore only $p_{\text{S}} = 0.25$, indicating hardly any evidence for
replication success.

<< "paradox-example" >>=
smdo <- 3
smdr <- -3
so <- sr <- 1
ps2 <- pSceptical(zo = smdo/so, zr = smdr/sr, c = so^2/sr^2,
                  alternative = "two.sided", type = "nominal")
@

There is one problem with the (two-sided) sceptical $p$-value as defined
in~\eqref{eq:ps}; It does not take the direction of the effect estimates into
account, so replication success may be established even though the effect
estimate from the replication goes in the opposite direction of the original one
($\sign(\that_{o}) \neq \sign(\that_{r})$). In fact, the sceptical $p$-value
will decrease with decreasing negative relative effect estimate
($\ps \downarrow p_{o}$ as $d \downarrow -\infty$ for fixed $\sigma_{o}$ and
$\sigma_{r}$) which seems contrary to intuitive understandings of replicability
as a change in effect direction may diminish the usefulness of a scientific
finding (\eg{} a negative treatment effect may indicate harm instead of
benefit).
% For example, if $\that_{o} = \Sexpr{smdo}$,
% $\that_{r} = \Sexpr{smdr}$, and $\sigma_{o} = \sigma_{r} = 1$, this would lead
% to $\ps = \Sexpr{pFormat(ps2)}$, and the sceptical $p$-value continues to
% decreases with decreasing replication effect estimates.
This ``replication
paradox'' \citep{Ly2018} can be avoided by defining a one-sided sceptical
$p$-value
\begin{align*}
  \tilde{p}_{\text{S}} =
  \begin{cases}
    \ps/2 & \text{if} ~\sign(\that_{o}) = \sign(\that_{r}) \\
    1 - \ps/2 & \text{if} ~ \sign(\that_{o}) \neq \sign(\that_{r}) \\
  \end{cases}
\end{align*}
similar to how ordinary one-sided $p$-values are defined. In the subsequent
chapters only the one-sided version of the sceptical $p$-value will be used.

>>>>

\begin{figure}[!htb]
<< "RPCB-graph", fig.height = 3.5 >>=
za <- qnorm(p = 0.975)
rpcb <- read.csv("data/rpcb.csv") |>
    mutate(id = paste0("(", paper, ", ", experiment, ", ", effect, ")"),
           zo = smdo/so,
           zr = smdr/sr,
           c = so^2/sr^2,
           d = smdr/smdo,
           po2 = 2*(1 - pnorm(q = abs(zo))),
           pr2 = 2*(1 - pnorm(q = abs(zr))),
           sm = 1/sqrt(1/so^2 + 1/sr^2),
           smdm = (smdo/so^2 + smdr/sr^2)*sm^2,
           pm2 = 2*(1 - pnorm(q = abs(smdm/sm))),
           Q = (smdo - smdr)^2/(so^2 + sr^2),
           pQ = pchisq(q = Q, df = 1, lower.tail = FALSE),
           rsig = factor(pr2 <= 0.05, levels = c(FALSE, TRUE),
                         labels = c("italic(p[r]) > 0.05",
                                    "italic(p[r]) <= 0.05")))
## compute sceptical p-value (TODO fix problems with NA)
ind <- !is.na(rpcb$zo) & !is.na(rpcb$zr) & !is.na(rpcb$c)
rpcb$ps1 <- NA
rpcb$ps1[ind] <- pSceptical(zo = rpcb$zo[ind], zr = rpcb$zr[ind], c = rpcb$c[ind],
                            alternative = "one.sided", type = "nominal")

@@

A detailed view for a subset of the data from the project is given in
Table~\ref{tab:RPCB}. The subset is chosen such that from every original study
one effect and one replication of that effect is included. For the three
replications at the bottom of the table (\#18, \#19, and \#20), all criteria
indicate replication failure. For the remaining majority of the studies, the
conclusions are more ambiguous; for no single replication study are all criteria
satisfied simultaneously. For instance, the first three studies satisfy the
significance criterion ($p_{o} < 0.05$ and $p_{r} < 0.05$), meta analytic
significance ($p_{m} < 0.05$) and the $Q$-test/prediction interval criterion
($p_{Q} > 0.05$), yet the replication effect estimate is smaller than the
original one ($d < 1$).
% For example, replication \#17 is a clear
% failure with respect to the relative effect estimate ($d < 1$) and the
% significance criteria ($p_{o} > 0.05$ and $p_{r} > 0.05$), yet the $Q$-test does
% not indicate evidence for inconsistency of the two effect estimates
% ($p_{Q} > 0.05$).

@@

\begin{landscape}
\begin{table}
  \centering
  \caption{Subset of results from the ``Reproducibility Project: Cancer
    Biology'' \citep{Errington2021}. (P, X, E) denotes effect number E from
    experiment number X, from original paper number P. Shown are original
    $\that_{o}$, replication $\that_{r}$, and pooled effect estimate $\that_{m}$
    with 95\% confidence intervals, variance ratio
    $c = \sigma^{2}_{o}/\sigma^{2}_{r}$, relative effect estimate
    $d = \that_{r}/\that_{o}$, original \textit{p}-value $p_{o}$, replication
    \textit{p}-value $p_{r}$, meta-analytic \textit{p}-value $p_{m}$, $Q$-test
    \textit{p}-value $p_{Q}$, and one-sided sceptical \textit{p}-value
    $p_{\mathrm{S}}$. A replication success (green) threshold of $\alpha = 5\%$
    is used for two-sided \textit{p}-values, $\alpha = 2.5\%$ for one-sided
    $p$-values, and $d_{\text{min}} = 1$ for the relative effect estimate.}
  \label{tab:RPCB}
\resizebox{1.02\linewidth}{!}{%
<< "RPCB-table2", results = "asis" >>=
## function to format CI
ciFormat. <- function(lower, upper, dollar = FALSE, digits = 2) {
    if (dollar) x <- "$" else x <- ""
    paste0("(", x, format(round(lower, digits), nsmall = digits), x,
           " to ", x, format(round(upper, digits), nsmall = digits), x, ")")
}
ciFormat <- Vectorize(FUN = ciFormat.)

@@

## randomly select a subset with one effect per paper
set.seed(7)
rpcbTable <- rpcb |>
    na.omit() |>
    group_by(paper) |>
    sample_n(size = 1) |>
    ungroup() |>
    arrange(ps1) |>
    mutate(smdoFormat = signif(smdo, 3),
           smdrFormat = signif(smdr, 3),
           smdmFormat = signif(smdm, 3),
           dFormat = tableFormat(x = d, outFun = dFormat,
                                 threshold = 1, col1 = "dark2orange",
                                 col2 = "dark2green"),
           cioL = smdo - za*so,
           cioU = smdo + za*so,
           cirL = smdr - za*sr,
           cirU = smdr + za*sr,
           cimL = smdm - za*sm,
           cimU = smdm + za*sm,,
           po2Format = tableFormat(x = po2),
           pr2Format = tableFormat(x = pr2),
           pm2Format = tableFormat(x = pm2),
           ps1Format = tableFormat(x = ps1, threshold = 0.025),
           pQFormat = tableFormat(x = pQ, col1 = "dark2orange",
                                  col2 = "dark2green"),
           cioFormat = ciFormat(lower = cioL, upper = cioU, dollar = TRUE),
           cirFormat = ciFormat(lower = cirL, upper = cirU, dollar = TRUE),
           cimFormat = ciFormat(lower = cimL, upper = cimU), dollar = TRUE) |>
    select(id,
           smdoFormat, cioFormat,
           smdrFormat, cirFormat,
           smdmFormat, cimFormat,
           c, dFormat,
           po2Format, pr2Format, pm2Format, pQFormat, ps1Format
           )
rpcbXtable <- xtable(rpcbTable)
colnames(rpcbXtable) <- c("(P, X, E)",
                          "$\\that_o$", "(95\\% CI)",
                          "$\\that_r$", "(95\\% CI)",
                          "$\\that_m$", "(95\\% CI)",
                          "$c$",
                          "$d$",
                          "$p_o$",
                          "$p_r$",
                          "$p_m$",
                          "$p_Q$",
                          "$p_{\\text{S}}$"
                          )

@@

The sceptical $p$-value has several interesting
properties \citep[Section 3]{Held2020}: First, it is always larger than the
maximum of the original and replication $p$-values
($p_{\text{S}} > \max\{p_{o}, p_{r}\}$), meaning that both $p$-values have to be
smaller than $\alpha$ such that replication success at level $\alpha$ is
possible. The sceptical $p$-value hence requires \emph{both} studies to be
sufficiently convincing on their own (in terms of their $p$-values), similar to
the significance criterion for replication success. Second, if the $p$-values
$p_{o}$ and $p_{r}$ remain fixed but the relative effect estimate
$d = \that_{r}/\that_{o}$ decreases towards zero, the sceptical $p$-value
increases, meaning that shrinkage of the replication effect estimate is
penalized ($p_{\text{S}} \uparrow 1$ as $d \downarrow 0$ for fixed $p_{o}$ and
$p_{r}$). The sceptical $p$-value thus takes effect shrinkage into account,
unlike the significance criterion which can be achieved with any non-zero
replication effect estimate $\that_{r}$ provided the standard error $\sigma_{r}$
is small enough. This property is desirable in the replication setting as a
smaller replication effect estimate $\that_{r}$ may not be practically relevant
anymore, despite its statistical significance.

There is one problem with the (two-sided) sceptical $p$-value as defined
in~\eqref{eq:ps}; It does not take the direction of the effect estimates into
account, so replication success may be established even though the effect
estimate from the replication goes in the opposite direction of the original one
($\sign(\that_{o}) \neq \sign(\that_{r})$). In fact, the sceptical $p$-value
will decrease with decreasing negative relative effect estimate
($\ps \downarrow p_{o}$ as $d \downarrow -\infty$ for fixed $\sigma_{o}$ and
$\sigma_{r}$) which seems contrary to intuitive understandings of replicability
as a change in effect direction may diminish the usefulness of a scientific
finding (\eg{} a negative treatment effect may indicate harm instead of
benefit).
% For example, if $\that_{o} = \Sexpr{smdo}$,
% $\that_{r} = \Sexpr{smdr}$, and $\sigma_{o} = \sigma_{r} = 1$, this would lead
% to $\ps = \Sexpr{pFormat(ps1)}$, and the sceptical $p$-value continues to
% decreases with decreasing replication effect estimates.
This ``replication
paradox'' \citep{Ly2018} can be avoided by defining a one-sided sceptical
$p$-value
\begin{align*}
  p_{\text{S}} =
  \begin{cases}
   1 - \Phi(|z_{\text{S}}|) & \text{if} ~\sign(\that_{o}) = \sign(\that_{r}) \\
   \Phi(|z_{\text{S}}|) & \text{if} ~ \sign(\that_{o}) \neq \sign(\that_{r}) \\
  \end{cases}
\end{align*}
similar to how ordinary one-sided $p$-values are defined. In the subsequent
chapters only the one-sided version of the sceptical $p$-value will be used.

The results from the ``Reproducibility Project: Cancer Biology'' in
Table~\ref{tab:RPCB} illustrate the properties of the sceptical $p$-value. For
example, study \#4 fails to achieve replication success at level
$\alpha = 2.5\%$ with the one-sided sceptical $p$-value, even though the
replication study was highly convincing (the effect estimate was almost five
times as large as in the original study). Yet, as the approach requires both
studies to be convincing on their own -- and the original study was not
significant at the 2.5\% level (one-sided) -- replication success at this level
is impossible with the sceptical $p$-value. The second property is illustrated
by study \#8. Here the original study was convincing and the replication study
also achieved significance. However, the effect estimate from the replication
was roughly 60\% smaller than the one from the original study, and significance
is merely achieved because the standard error was much smaller (\ie{} around
$1/\surd c \approx 1/2.8$ times smaller). The sceptical $p$-value is therefore
only $p_{\text{S}} = 0.12$, indicating hardly any evidence for replication
success.

<< "paradox-example" >>=
smdo <- 3
smdr <- -3
so <- sr <- 1
ps1 <- pSceptical(zo = smdo/so, zr = smdr/sr, c = so^2/sr^2,
                  alternative = "two.sided", type = "nominal")
@
